Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-08-17 Manuscript/manuscript.tex
added appendix

\newcommand{\EHF}{E_\text{HF}}
\newcommand{\EDOCI}{E_\text{DOCI}}
\newcommand{\EFCI}{E_\text{FCI}}
\newcommand{\Ndet}{N_\text{det}}
\newcommand{\Nbas}{N}

@@

Finally, one can employ a reference determinant having more unpaired electrons to define the hCI models.
The case of two unpaired electrons ($s_0=2$) is shown in the right pannel of Fig.~\ref{fig:hCI},
which would be appropriate for describing singly-excited states of closed-shells.
The key difference in the hCI sequence, with respect to the closed-shell case, lies in the displacement by one block to the right in the seniority-excitation map,
reflecting the shift from the $s_0=0$ to the $s_0=2$ reference.
In addition, in the two previous cases, the hCI0 level only included the reference determinant.
whereas for $s_0=2$, hCI0 not only accounts for the reference determinant,
but also for the two closed-shell determinants produced by the single excitation that pairs the two unpaired electrons.
The hCI1 classes of determinants for the $s_0=2$ reference can be seen in the bottom of Fig.~\ref{fig:determinants},
which clearly outnumber the fewer classes associated with the $s_0=0$ and $s_0=1$ cases.
One could adopt references with even larger $s_0$, further displacing the hCI sequence to the right in the seniority-excitation map.
In this case, hCI0 would include all the lower spin determinants obtained by partially or totally pairing the unpaired electrons.
Notice that the number of such determinants do not depend on the size of the system (scaling as $N^0$), and are thus included at the $h=0$ level, in line with the spirit of hCI.
In a similar fashion, hCI can be built on top of a larger $s_0$ reference for odd number of electrons.

@@

>>>>

\newcommand{\Na}{N_{\alpha}}
\newcommand{\Nb}{N_{\beta}}
\newcommand{\Nd}{N^{d}}
\newcommand{\Nv}{N^{v}}
\newcommand{\Nas}{N^s_{\alpha}}
\newcommand{\Nbs}{N^s_{\beta}}

\newcommand{\EHF}{E_\text{HF}}
\newcommand{\EFCI}{E_\text{FCI}}
\newcommand{\Ndet}{N_\text{det}}
\newcommand{\Nbas}{N}

@@

Finally, one can employ a reference determinant having more unpaired electrons to define the hCI models.
The case of two unpaired electrons ($s_0=2$) is shown in the right pannel of Fig.~\ref{fig:hCI},
which would be appropriate for describing singly-excited states of closed-shells.
The key difference in the hCI sequence, with respect to the closed-shell case, lies in the displacement by one block to the right in the seniority-excitation map,
reflecting the shift from the $s_0=0$ to the $s_0=2$ reference.
In addition, in the two previous cases, the hCI0 level only included the reference determinant.
whereas for $s_0=2$, hCI0 not only accounts for the reference determinant,
but also for the two closed-shell determinants produced by the single excitation that pairs the two unpaired electrons.
The hCI1 classes of determinants for the $s_0=2$ reference can be seen in the bottom of Fig.~\ref{fig:determinants},
which clearly outnumber the fewer classes associated with the $s_0=0$ and $s_0=1$ cases.
One could adopt references with even larger $s_0$, further displacing the hCI sequence to the right in the seniority-excitation map.
In this case, hCI0 would include all the lower spin determinants obtained by partially or totally pairing the unpaired electrons.
Notice that the number of such determinants do not depend on the size of the system (scaling as $N^0$), and are thus included at the $h=0$ level, in line with the spirit of hCI.
In a similar fashion, hCI can be built on top of a larger $s_0$ reference for odd number of electrons.
The deduction for the number of determinants in a given hCI model, for the $s_0=0$ closed-shell, $s_0=1$ open-shell, as well as for an arbitrary reference determinant,
can be found in Sec.~\ref{sec:appendix}.

@@

\appendix

\section{Number of determinants}
\label{sec:appendix}

What is the number of determinants in a given hCI model, defined by the hierarchy $h$ [see Eq.~\eqref{eq:h}] and the reference determinant?
Here, we address this question by first working out the simpler case of a closed-shell reference.
Then, we move to a slightly more complicated case of an open-shell reference with a single unpaired electron.
Finally, we deduce the general case.
The reference determinant of each case is shown in Fig.~\ref{fig:appendix}.

%%% FIG 5 %%%
\begin{figure}[h!]
\includegraphics[width=1.0\linewidth]{determinants_appendix}
\caption{
$s_0=0$ closed-shell determinant with $N$ electrons (left),
$s_0=1$ open-shell determinant with $\Na$ $\alpha$-electrons and $\Nb=\Na-1$ $\beta$-electrons (center),
and general $s_0=\Nas+\Nbs$ determinant, with $\Nd$ doubly-occupied orbitals, $\Nas$ singly-occupied $\alpha$-electrons and $\Nbs$ singly-occupied $\beta$-electrons (right).
In the three cases, there are $K$ spatial orbitals.
The number of determinants generated from these three reference determinants, from left to right, which have excitation degree $e$ and seniority $s$
are given in Eqs.~\eqref{eq:det_1}, ~\eqref{eq:det_2}, and ~\eqref{eq:det_3}.
}
\label{fig:appendix}
\end{figure}
%%% %%% %%%

The gist of the deduction is as follows.
We start from the known number of determinants based only on the excitation degree $e$.
Then, we systematically decompose each term into specific contributions,
based on whether the excitation increases or reduces the seniority.
For the decomposition, we make use of the Vandermonde's identity,
\begin{equation}
	\binom{n}{k} = \sum_{j}^{k} \binom{m}{j} \binom{n-m}{k-j}.
\end{equation}
Next, an incremental seniority can be assigned to each type of contribution.
Finally, the final answer can be obtained by summing only the contributions with the desired seniority $s$.

We start with the simpler case of a closed-shell determinant with $N$ electrons and $K$ spatial orbitals (left of Fig.~\ref{fig:appendix}).
From this reference determinant, the number of excited determinants generated by exciting exactly $e$ electrons, is given by \cite{SzaboBook}
\begin{equation}
\sum_{p=0}^{e} \binom{N/2}{p} \binom{K-N/2}{p} \binom{N/2}{e-p} \binom{K-N/2}{e-p},
\label{eq:e_closed}
\end{equation}
where the sum expresses the different combinations for exciting $\alpha$- or $\beta$-electrons,
the first binomial term accounts for the number of possibilities for exciting $p$ $\alpha$-electrons from the $N/2$ orbitals,
the second term represents the number of ways of placing these $p$ electrons into $K-N/2$ orbitals,
and similarly for the latter two terms and the $e-p$ remaining $\beta$-electrons.

To account for the seniority of the excited determinants, one must disentangle the excitations based on how they change the seniority number,
while keeping track of the previously excited electrons.
Starting from the closed-shell determinant, $p$ $\alpha$-electrons are excited, which increases the seniority by $2p$
(a factor of $p$ due to the unpaired $\alpha$-electrons just excited,
and another factor of $p$ due to the unpaired $\beta$-electrons left behind).
The first two binomial terms in Eq.~\ref{eq:e_closed} are left untouched since they are always accompanied by the same change in seniority.
Next, the $\beta$-electrons can be excited from the same orbitals from which the $\alpha$-electrons were excited (decreasing the seniority by one)
or instead from an orbital that remained doubly-occupied (increasing the seniority by one).
The two possibilities are expressed by decomposing the third binomial term in Eq.~\ref{eq:e_closed} as a sum over the two corresponding binomials,
\begin{equation}
\binom{N/2}{e-p} = \sum_{q=0}^{e-p} \binom{p}{q} \binom{N/2-p}{e-p-q}.
\end{equation}
The first counts the number of possibilities for exciting $q$ $\beta$-electrons from one of $p$ orbitals for which an $\alpha$-electron was already excited.
By removing the unpaired $\beta$-electron left behind, the seniority thus decreases by a factor of $q$.
The second term accounts for the complementary excitations, where the $(e-p-q)$ $\beta$-electrons are chosen out of the $N/2-p$ orbitals that 
remained doubly-occupied after exciting the $p$ $\alpha$-electrons.
Therefore, this term increases the seniority by $(e-p-q)$.
We proceed similarly for the fourth binomial term of Eq.~\ref{eq:e_closed}, by decomposing it into the sum of two other binomials,
where $r$ $\beta$-electrons pair with the previously $\alpha$-electrons (thus reducing the seniority) and $(e-p-r)$ do not (which increases the seniority).
By collecting the seniority changes from each term: $(2p)$ from the unmodified binomials, $(-q)$ and $(e-p-q)$ from the third binomial, $(-r)$ and $(e-p-r)$ from the fourth one,
the seniority of the excited determinant is given by $s = 2 (e-q-r)$.
Finally, by combining the seniority specific binomials and imposing the desired seniority via a Kronecker delta,
the number of determinants with a given excitation degree $e$ and seniority $s$ produced from a closed-shell reference ($s_0=0$) is given as
\begin{equation}
\begin{split}
	&\sum_{p=0}^{e} \binom{N/2}{p} \binom{K-N/2}{p}
	\sum_{q=0}^{e-p} \binom{p}{q} \binom{N/2-p}{e-p-q}  \\
	&\sum_{r=0}^{e-p} \binom{p}{r} \binom{K-N/2-p}{e-p-r}
 	\delta_{s,2(e-q-r)}.
\label{eq:det_1}
\end{split}
\end{equation}
To obtain the final number of determinants for a given hierarchy $h$, we simply sum over the allowed combinations of $e$ and $s$ according to Eq.~\eqref{eq:h}.

Moving to the case where the number of $\alpha$-electrons ($\Na$) and $\beta$-electrons ($\Nb$) differ by one ($\Na=\Nb+1$), which is illustrated in the center of Fig.~\ref{fig:appendix}.
The number of excited determinants generated exclusively by exciting $e$ electrons %$\Ndet(e)$, 
is analogous to Eq.~\eqref{eq:e_closed}, being given by
\begin{equation}
\sum_{p=0}^{e} 
\binom{\Na}{p} \binom{K-\Na}{p} \binom{\Nb}{e-p} \binom{K-\Nb}{e-p}.
\label{eq:e_open}
\end{equation}

The deduction proceeds similarly to the closed-shell case. 
The main differences are the following.
The first binomial term of Eq.~\ref{eq:e_open} is decomposed into the sum of $t$ $\alpha$-electrons excited from the singly-occupied orbital and $(p-t)$ that are excited from the $\Nb$ doubly-occupied ones.
The second binomial is left untouched, as before.
The third one is also decomposed as explained for the closed-shell case, with the difference that the $q$ $\beta$-electrons are excited from $(p-t)$ orbitals (instead of $p$) 
from which $\alpha$-electrons were excited. 
Similarly, in the fourth binomial of Eq.~\ref{eq:e_open}, $r$ $\beta$-electrons are placed into one of the $(1+p-t)$ orbitals (rather than $p$ in the closed-shell case) which contain an unpaired $\alpha$-electron.
Collecting the individual contributions and setting the targeted seniority,
the number of determinants generated from a $s_0=1$ reference determinant, with excitation degree $e$ and seniority $s$ is given by
\begin{equation}
\begin{split}
	& \sum_{p=0}^{e} 
	\sum_{t=0}^{p} \binom{1}{t} \binom{\Nb}{p-t} 
	\binom{K-\Na}{p} 
	\sum_{q=0}^{e-p} \binom{p-t}{q} \binom{\Nb-p+t}{e-p-q} \\
	& \sum_{r=0}^{e-p} \binom{1+p-t}{r} \binom{K-\Nb-1-p+t}{e-p-r}
 	\delta_{s,1+2(e-q-r-t)}.
\label{eq:det_2}
\end{split}
\end{equation}

In the general case, an arbitrary reference determinant is defining by three parameters, the number of doubly-occupied orbitals ($\Nd$),
the number of singly-occupied $\alpha$-electrons ($\Nas = \Na - \Nd$),
and the number of singly-occupied $\beta$-electrons ($\Nbs = \Nb - \Nd$).
(Alternatively, one could employ the number of electrons $N$, $\Nd$, and the spin quantum number $S_z = (\Na-\Nb)/2$.)
For $K$ spatial orbitals, the number of virtual orbitals is $\Nv = K - \Nd - \Nas - \Nbs$.

The deduction is somewhat more involved, but follows along the same lines as the two previously discussed cases.
The starting point is also given by Eq.~\ref{eq:e_open}, with each binomial being written as a sum of two or more binomials that distinguish the change in seniority.
Here, we just state how each term is decomposed, where the details can be checked by inspection of the final result presented below [Eq.~\eqref{eq:det_3}].
The first binomial of Eq.~\ref{eq:e_open} is decomposed into the contributions of $\alpha$-electrons excited from the $\Nd$ doubly- and from the $\Nas$ singly-occupied orbitals.
The second binomial is decomposed into the $\alpha$-electrons being excited to the $\Nv$ virtual orbitals and to the $\Nbs$ singly-occupied orbitals.
The third one undergoes two decompositions, and is thus expressed as a double sum over three binomials.
The $\beta$-electrons are distinguished between the $\Nbs$ singly-occupied orbitals, a subset of $\Nd$ doubly-occupied orbitals for which $\alpha$-electrons were excited,
and the complementary subset for which no $\alpha$-electrons were excited.
Finally, the fourth binomial of Eq.~\ref{eq:e_open} counts the number of possibilities to place the excited $\beta$-electrons.
It is first decomposed into two terms, based on whether they are promoted to the $\Nv$ virtual orbitals or to the $\Nas$ singly-occupied orbitals.
Each term is further decomposed into two terms, accounting for the subset of $\Nv$ orbitals which now contain $\alpha$-electrons,
and the subset of $\Nas$ orbitals from which $\alpha$-electrons were removed,
thus leading to three sums over four binomials.
The final result for the number of determinants generated from an arbitrary reference determinant, constrained to have excitation degree $e$ and seniority $s$ is given by
\begin{equation}
\begin{split}
	& \sum_{p=0}^{e} 
	\sum_{t=0}^{p} \binom{\Nas}{t} \binom{\Nd}{p-t} 
	\sum_{w=0}^{p} \binom{\Nv}{w} \binom{\Nbs}{p-w} \\
	& \sum_{u=0}^{e-p} \binom{\Nbs}{u} 
	\sum_{v=0}^{e-p-u} \binom{p-t}{v} \binom{\Nd-p+t}{e-p-u-v} \\
	& \sum_{r=0}^{e-p} \sum_{m=0}^{r} \binom{w}{m} \binom{\Nv-w}{r-m} 
	\sum_{n=0}^{e-p-r} \binom{t}{n} \binom{\Nas-t}{e-p-r-n} \\
	& \delta_{s,\Nas+\Nbs+2(e+w-p-t-u-v-m-n)}.
\label{eq:det_3}
\end{split}
\end{equation}
